### 2024-06-10 Paper/paper.tex
Updates from Overleaf

    \begin{abstract}
		{\small
        In many industries, not-for-profits (NFP) are the dominant firm type. Understanding the objectives of these firms is important, but NFP firms have underlying goals that make understanding their objectives difficult. For hospitals in particular, there is mixed evidence on whether NFP hospitals differ meaningfully from for-profit (FP) hospitals. In this paper, I investigate whether leadership team composition of NFP hospitals affects how similarly they behave to for-profits. I use a novel data set on NFP hospital executives in the US (2010-2014) to investigate whether the presence of clinically trained executives affects hospital behavior. In particular, I leverage pay-for-performance initiatives in the US enacted in 2012, and measure the  readmission and mortality rate responses between FP, NFP, and NFP with and without clinically trained executives. I find that NFPs without clinically trained executives behave more like FPs than other NFPs. This is consistent with theoretical predictions if NFP hospitals with clinically trained executives value societal outcomes more than those without clinically trained executives.
		} 
	\end{abstract}

@@

  The objectives and behaviors of firms that are not classic for-profits are not easily determined, and an extensive literature has sought to understand these types of firms. Not-for-profits (NFP) are thought to gain utility not solely from monetary profit, but also through accomplishing societal benefit. Researchers have speculated a variety of objective functions that could motivate NFP firms such as maximizing profit (for-profit in disguise), prestige, income, or a quality/quantity trade-off (\cite{steinberg1986revealed}). Empirical contributions to this literature center around investigating behaviors that differ between for-profit (FP), private NFP, and government firms (\cite{sloan2000not}). In this paper, I investigate whether differences in observed behavior between for-profit and not-for-profit hospitals vary differentially by leadership team composition.

  Hospitals have been a common focus when seeking to understand NFP firms. This is because, first, there are different hospital ownership types within the industry, giving researchers the ability to compare the behaviors of NFPs and FPs. Second, private NFP hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, FPs make up around 36\% of hospitals and staff 107 beds (\cite{ASPE_2023}). Thus, NFP behavior in the hospital industry is relevant, and directly affects consumers of health care. However, it remains unclear what objective function drives NFP hospital behavior, and whether NFP hospitals are meaningfully different from FP hospitals (\cite{sloan2000not}; \cite{erus2002inferring}; \cite{deneffe2002not}; \cite{horwitz2009hospital}).

  While ownership is certainly relevant, there could be other characteristics correlated with hospital behavior independent of ownership type. Researchers have focused in particular on explaining variation in NFP behavior using variation in the market they exist in (\cite{horwitz2009hospital}). However, I argue that an observable and economically meaningful factor is the composition of top-level hospital managers. In large, publicly traded firms, executive characteristics are shown to be correlated with firm performance, indicating that there is more to firm behavior than ownership status (\cite{bertrand2003managing}; \cite{matsa2013female}; \cite{ahern2012changing}). If hospitals differ in how they value monetary profit vs. societal benefit, then ownership status certainly plays a role in these valuations, but the question remains as to whether management also plays an important role. I contribute to our understanding of hospital and NFP behaviors by investigating whether occupational background, specifically clinical training in executives, affects hospital response to pay-for-performance incentives.

@@

  I estimate how the interaction between hospital type and post program enactment affects readmission and mortality rates. Under the assumption that hospitals would have continued to behave similarly in the absence of the programs, and that leadership teams do not change endogenously with the programs, I identify the differential response to pay-for-performance incentives among hospital types. That is, I define a group of treatment and comparison hospitals, and estimate how outcomes differ between the two groups after the policy changes. I ultimately employ synthetic difference-in-differences to analyze a group of comparison hospitals with similar characteristics to treatment hospitals.\footnote{I also present results for the full sample, and find qualitatively similar results.}

   While all hospital types decrease readmission rates after the programs are enacted, NFPs with clinically trained executives respond less than both FPs and NFPs without clinically trained executives. That is, NFPs without clinically trained executives respond to incentives more similarly to FPs than other NFPs. Leveraging changes in executive teams, I decompose this effect into signalling (clinical executives signal underlying hospital preferences), or managing (clinical executives manage the hospital differently) effects. I find that the entire difference in response is driven by a managing effect.

    This paper contributes to three strands of literature. First, it adds to our understanding of NFP hospital objective functions. There is mixed evidence on whether NFP behavior differs from FP behavior: costs, uncompensated care, technology adoption, and quality (\cite{sloan2000not}; \cite{eggleston2008hospital}; \cite{moscelli2018effect}; \cite{moscone2020public}). To this point, researchers have found that NFPs do not act as purely profit maximizers, but maximize some combination of profit and output (\cite{deneffe2002not}, \cite{chang2011nonprofit}). A paper with an approach similar to mine which investigates hospital behavior is \citeauthor{chang2011nonprofit} (\citeyear{chang2011nonprofit}), which looks at the effect of a cost shock on likelihood of shutting down and mix of profitable services. They find that NFPs and FPs are equally as likely to shut down after the shock, but only NFPs adjust their mix of profitable services. Similarly, I investigate a change in incentives to hospitals, but I look at the differential impacts of leadership teams within not-for-profits. This paper adds to our understanding of inputs into not-for-profit hospital objective functions.

@@

    Not-for-profit firms in the US differ from for-profits mainly in that for-profit firms distribute revenue to stakeholders, while not-for-profits reinvest profit back into the firm. An important distinction between the two is that NFPs are, in general, driven by goals to better society, and therefore receive tax benefits from the government.

    Hospitals are governed by a board of directors, whose role is to set broad goals and strategies, and provide general oversight. The board selects executives, the highest level of management of a firm, to carry out day-to-day operations. An executive team usually consists of at least a Chief Executive Officer (CEO) and Chief Financial Officer (CFO), but there is variation in how firms organize executive teams. It is not uncommon for hospital executives to specialize in health care settings by earning a degree in health care management, or an MBA specific to health care. There are executive positions that are often filled by someone with clinical training, such as a Chief Medical Officer (CMO), and doctors can even find themselves in other top managerial positions as well, such as CEO or president. While some doctors earn additional degrees before stepping into an executive role, this is not a necessary condition to becoming a physician executive.

@@

    In October 2011, the Center for Medicare and Medicaid Services (CMS) released a set of rules under HRRP mandating penalties for hospitals with above average readmission rates. A readmission is when a patient returns to the hospital within 30 days of being discharged from a previous stay; avoidable readmissions are a bad outcome for patients and increase health care spending. The goal of HRRP is to lower readmissions through better care coordination, less initial stay complications, and better post-care instructions. Beginning in October 2012, hospitals with higher readmission rates than the national average in pneumonia, heart failure, or AMI (after adjusting for demographic characteristics) receive a fixed lower reimbursement rate for all Medicare patients seen in their hospital. In 2015, CMS also included chronic obstructive pulmonary disease, coronary artery bypass graft surgery, and elective primary total hip arthroplasty and/or total knee arthroplasty as conditions which go into the penalty calculation (\cite{CMS}).

    Penalties are given in the form of a fixed rate reduction of 1-3\% for every Medicare patient regardless of the condition. Further, CMS does not distinguish a necessary readmission from an avoidable readmission; any repeat hospital visit is included in the penalty calculation. Excess readmission rates are calculated using a rolling look-back period of 3 years to determine whether the hospital is penalized. Therefore, hospitals had incentive to react immediately once details of the program were announced in October of 2011. These penalties are not insignificant; penalized hospitals paid, on average, 4-5\% of revenue.

@@

    For each of the matched EINs, I extract web URLs to Form 990 PDFs for each year. I use an algorithm to download these locally and use optical character recognition (OCR) text extraction methods to scrape the relevant section of the PDF, which is titled ``Officers, Directors, Trustees, Key Employees, and Highest Compensated Employees". Finally, I use string cleaning methods to identify all names and positions under each EIN, year. With each name, I extract titles indicating clinical training: MD, Dr., or DO. Using OCR extraction can be unreliable in cases where the text on the form is handwritten or small. Therefore, for observations that are missing after the initial extraction, I manually record names, positions, and titles. This yields 852 not-for-profit hospitals that are matched to AHA data, and are not missing text information.

    For the purposes of this paper, I focus mainly on executive teams as a whole instead of specific positions. The main reason for this is because the organization of teams varies so drastically across hospitals that a specific position, even if it exists in two different hospitals, can be meaningfully different. Additionally, the more granular the information I extract, the more likely there will be a mistake in the OCR extraction. Thus, to preserve correctness of the independent variable of interest, I focus on whether there is any MD executive at the hospital.

    I drop anyone who is solely a board member, and focus on those who have executive in their title, or are labeled as a president or vice president of a department. I create various hospital-level characterizations based on their executive team, such as the existence of a clinically trained executive, the number of clinically trained executives, the number of total executives, and whether the hospital has a Chief Medical Officer. Additionally, I examine changes in propensity to hire a clinical executive. Summary statistics for these variables are shown in Table \ref{leader_sumstats}. On average, NFP hospitals in my sample have 4 executives and .4 clinically trained executives. Sixty-one percent of the sample never changes propensity to hire a clinically trained executive, and 33\% of NFPs in the sample have a Chief Medical Officer at some point.

@@

    I combine the created sample of not-for-profits with all for-profits in the AHA survey. The outcomes I consider are readmission and mortality rates, which are directly targeted in the pay-for-performance policies, and are found in the CMS Hospital Compare data. I combine rates for pneumonia, AMI, and heart failure (the relevant HRRP conditions) into a weighted average based on the number of patients in each condition. Additionally, I include several relevant hospital characteristics: number of beds, whether the hospital is an academic medical center, is physician owned, or is system affiliated, all come from the AHA survey. Case mix index, a measure of overall patient complexity, comes from CMS. Whether the hospital was penalized under HRRP or received payments under HVBP is found in the HCRIS data. Finally, I limit to years 2010-2014.

@@

@@

    \section{Model and Empirical Strategy}\label{sec:model}

@@

    \subsection{Estimation}

@@

@@

    \section{Results}

    I also directly estimate the difference in mortality rates among NFPs with and without clinical experience in panel \ref{fig:mort_synth_plotd}. Again, the difference is not statistically significant, but the coefficient compared to the true nulls in panels \ref{fig:mort_synth_plota} and \ref{fig:mort_synth_plotc} indicates NFPs without a clinical executive potentially decrease mortality more aggressively than NFPs with a clinical executive.

    \subsection{Decomposition}

@@

\section{Results By Condition}

@@

@@

>>>>

    \begin{abstract}
		{\small
        There is mixed evidence on whether not-for-profit hospitals differ meaningfully from for-profit (FP) hospitals, but what has not been considered is the effect leadership teams have on hospital behavior. In this paper, I investigate whether leadership team composition of NFP hospitals affects how similarly they behave to for-profits. I use a novel data set on NFP hospital executives in the US (2010-2014) to investigate whether the presence of clinically trained executives affects hospital behavior. In particular, I leverage pay-for-performance initiatives in the US enacted in 2012, and measure the  readmission and mortality rate responses between FP and NFP with and without clinically trained executives. I find that for-profits nad NFPs without clinical executives respond more aggressively to financial incentives on quality, consistent with theoretical predictions of firms that value profit relatively more than having high quality. 
		} 
	\end{abstract}

@@

  The objectives and behaviors of not-for-profit firms are challenging to determine, and an extensive literature has sought to understand them. Researchers have posited a variety of objective functions that could motivate NFP firms such as maximizing profit (for-profit in disguise), prestige, income, or a quality/quantity trade-off (\cite{steinberg1986revealed}). Empirical contributions to this literature center around investigating behaviors that differ between for-profit (FP), private NFP, and government firms (\cite{sloan2000not}). However, the literature has neglected to consider other relevant characteristics of hospitals that could affect objectives, such as the hospital's leadership team. In this paper, I investigate whether differences in observed behavior between for-profit and not-for-profit hospitals vary differentially by leadership team composition.

  Hospitals have been a common focus when seeking to understand NFP firms. This is because, first, there are different hospital ownership types within the industry, giving researchers the ability to compare the behaviors of NFPs and FPs. Second, private NFP hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, FPs make up 36\% of hospitals and staff 107 beds (\cite{ASPE_2023}). Thus, NFP behavior in the hospital industry is relevant, and directly affects consumers of health care. However, it remains unclear what objective function drives NFP hospital behavior, and whether NFP hospitals are meaningfully different from FP hospitals (\cite{sloan2000not}; \cite{erus2002inferring}; \cite{deneffe2002not}; \cite{horwitz2009hospital}).

  While ownership is certainly relevant, there could be other characteristics correlated with hospital behavior independent of ownership type. Researchers have focused in particular on explaining variation in NFP behavior using variation in the market they exist in (\cite{duggan2002hospital}; \cite{horwitz2009hospital}). However, I argue that an observable and economically meaningful factor is the composition of top-level hospital managers. In large publicly traded firms, executive characteristics are shown to be correlated with firm performance and governance, indicating that there is more to firm behavior than ownership status (\cite{bertrand2003managing}; \cite{ahern2012changing}\cite{matsa2013female}).

  A policy relevant aspect of hospital behavior in the US has been the shift towards pay-for-performance incentives. While quality may or may not directly matter to hospitals, pay-for-performance incentives are such that quality of care directly impacts profit. For hospitals that already valued quality before pay-for-performance incentives, the change in incentives may not change quality much. However, if the hospital is more profit driven, the policy change should translate to a larger change in quality. Ownership status certainly plays a role in the valuations of profit vs. quality, but the question remains as to whether management also plays an important role. I contribute to our understanding of hospital and NFP behaviors by investigating whether occupational background of executives, specifically clinical training, affects hospital response to pay-for-performance incentives.

@@

  I estimate how the interaction between hospital type and pay-for-performance incentives affects readmission and mortality rates. Under the assumption that hospitals would have continued to behave similarly in the absence of the programs, and that leadership teams do not change endogenously with the programs, I identify the differential response to pay-for-performance incentives among hospital types. That is, I define a group of treatment and comparison hospitals, and estimate how outcomes differ between the two groups after the policy changes. I ultimately employ synthetic difference-in-differences to analyze a group of comparison hospitals with similar characteristics to treatment hospitals.\footnote{I also present results for the full sample and find qualitatively similar results.}

   While quality generally improves after pay-for-performance incentives for all hospital types, the most aggressive response is from for-profits and NFPs without clinical executives, who decrease readmission and mortality rates more so than NFPs with clinical executives. That is, the objectives of NFPs without clinically trained executives seem to align more with those of for-profits: placing more weight on profit than societal outcomes. Leveraging changes in executive teams, I decompose this effect into signalling (clinical executives signal underlying hospital preferences), or managing (clinical executives manage the hospital differently) effects. I find that the entire difference in response is driven by a managing effect.

    This paper contributes to three strands of literature. First, it adds to our understanding of NFP hospital objective functions. There is mixed evidence on whether NFP behavior differs from FP behavior in terms of costs, uncompensated care, technology adoption, and quality (\cite{sloan2000not}; \cite{eggleston2008hospital}; \cite{moscelli2018effect}; \cite{moscone2020public}). To this point, researchers have found that NFPs do not act as purely profit maximizers, but maximize some combination of profit and output (\cite{deneffe2002not}, \cite{chang2011nonprofit}). This paper adds to our understanding of inputs into not-for-profit hospital objective functions.

@@

    Not-for-profit firms in the US differ from for-profits mainly in that for-profit firms distribute revenue to stakeholders, while not-for-profits reinvest profit back into the firm. Additionally, NFP firms are generally expected to set goals to better society due to government tax breaks received.

    Hospitals are governed by a board of directors, whose role is to set broad goals and strategies, and provide general oversight. The board selects executives, the highest level of management of a firm, to carry out day-to-day operations. An executive team usually consists of at least a Chief Executive Officer (CEO) and Chief Financial Officer (CFO), but there is variation in how firms organize executive teams. It is not uncommon for hospital executives to specialize in health care settings by earning a degree in health care management, or an MBA specific to health care. There are executive positions that are often filled by someone with clinical training, such as a Chief Medical Officer (CMO), and medical doctors pursue other top managerial positions as well, such as CEO or president. While some doctors earn additional degrees before stepping into an executive role, this is not a necessary condition to becoming a physician executive.

@@

    In October 2011, the Center for Medicare and Medicaid Services (CMS) released a set of rules under HRRP mandating penalties for hospitals with above average readmission rates. The goal of HRRP is to lower readmissions through better care coordination, less initial stay complications, and better post-care instructions. Beginning in October 2012, hospitals with higher readmission rates than the national average in pneumonia, heart failure, or AMI (after adjusting for demographic characteristics) receive a fixed lower reimbursement rate for all Medicare patients seen in their hospital. In 2015, CMS also included chronic obstructive pulmonary disease, coronary artery bypass graft surgery, and elective primary total hip arthroplasty and/or total knee arthroplasty as conditions which go into the penalty calculation (\cite{CMS}).

    Penalties are given in the form of a fixed rate reduction of 1-3\% for every Medicare patient regardless of the condition. Further, CMS does not distinguish a necessary readmission from an avoidable readmission; any repeat hospital visit is included in the penalty calculation. Excess readmission rates are calculated using a rolling look-back period of 3 years to determine whether the hospital is penalized. Therefore, hospitals had incentive to react immediately once details of the program were announced in October of 2011.

@@

    For each of the matched EINs, I extract web URLs to Form 990 PDFs for each year. I use an algorithm to download these locally and use optical character recognition (OCR) text extraction methods to scrape the relevant section of the PDF, which is titled ``Officers, Directors, Trustees, Key Employees, and Highest Compensated Employees". Finally, I use string cleaning methods to identify all names and positions under each EIN, year. With each name, I extract titles indicating clinical training: MD, Dr., or DO. Using OCR extraction can be unreliable in cases where the text on the form is handwritten or small. Therefore, for observations that are missing after the initial extraction, I manually record names, positions, and titles. As I limit the sample to hospitals who have text information in every year, I limit the years to 2010-2014 to retain more hospitals. This yields 852 not-for-profit hospitals that are matched to AHA data, and with complete text information in each year.

    For the purposes of this paper, I focus mainly on executive teams as a whole instead of specific positions. The main reason for this is because the organization of teams varies so drastically across hospitals that a specific position, even if it exists in two different hospitals, can be meaningfully different. I drop anyone who is solely a board member, and focus on those who have executive in their title, or are labeled as a president or vice president of a department. I create various hospital-level characterizations based on their executive team, such as the existence of a clinically trained executive, the number of clinically trained executives, the number of total executives, and whether the hospital has a Chief Medical Officer. Additionally, I examine changes in propensity to hire a clinical executive. Summary statistics for these variables are shown in Table \ref{leader_sumstats}. On average, NFP hospitals in my sample have 4 executives and .4 clinically trained executives. Sixty-one percent of the sample never changes propensity to hire a clinically trained executive, and 33\% of NFPs in the sample have a Chief Medical Officer at some point.

@@

    I combine the resulting sample of not-for-profits with all for-profits in the AHA survey. The outcomes I consider are readmission and mortality rates, which are directly targeted in the pay-for-performance policies, and are found in the CMS Hospital Compare data. I combine rates for pneumonia, AMI, and heart failure (the relevant HRRP conditions) into a weighted average based on the number of patients in each condition. Additionally, I include several relevant hospital characteristics from the AHA survey: number of beds, and whether the hospital is an academic medical center, is physician owned, or is system affiliated. Case mix index, a measure of overall patient complexity, comes from the CMS Case Mix Index files. Whether the hospital was penalized under HRRP or received payments under HVBP is found in the HCRIS data.

@@

    In Table \ref{tab:sumstats_samples}, I present a table of means of relevant variables for several hospital samples used throughout the analysis. First, there are 708 for-profit hospitals in the sample, shown in the left-most column. Notably,

@@

    \section{Modelling Hospital Response to Pay-for-Performance}\label{sec:model}

@@

    \section{Estimating Differences in Hospital Response}

    \subsection{Method}

@@

 \subsection{Identification}

 \subsection{Results}

    I also directly estimate the difference in mortality rates among NFPs with and without clinical experience in panel \ref{fig:mort_synth_plotd}. Again, the difference is not statistically significant, but the coefficient compared to the true nulls in panels \ref{fig:mort_synth_plota} and \ref{fig:mort_synth_plotc} indicates NFPs without a clinical executive potentially decrease mortality more aggressively than NFPs with a clinical executive. In Appendix \ref{app:condition}, I show that these results are mainly driven by pneumonia patients.

\section{Signalling vs. Managing Decomposition}

\subsection{Method}

@@

    \subsection{Results}

@@

\section{Results By Condition}\label{app:condition}

The main results of the paper consider the weighted average readmission or mortality rate as the main outcomes. Here, I present results for each condition individually. Table \ref{tab:fp_read_condition_synth.tex}, I show the effect of being for-profit on AMI, heart failure, and pneumonia readmission rates compared to differing NFP comparison groups. The estimates indicate that the weighted average is mainly driven by pneumonia patients, since AMI readmission rate responses differ between FP and all NFP, where no other condition does.

@@

Table \ref{tab:fp_mort_condition_synth.tex} shows a very similar result for mortality rates. AMI responses differ between FP and all NFP or NFP without clinical executives, but the weighted average is driven mainly by pneumonia patients.

@@

The differences between NFPs with and without clinical executives are also driven mainly by pneumonia patients, shown in Table \ref{tab:MD_noMD_readmort_condition_synth}.
